A formal treatment of bidirectional typing

### 2024-01-12 tex/ESOP24/S1-intro.tex
[ tex ] Eliminate most of hyphens the end of every line

As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} to explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

@@

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a \emph{syntax-directed} bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
For instance, we briefly discuss how the theory could be extended to handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed~\cite{Chen2024a} with \Agda but translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, with no gap in between.

>>>>

As we mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} to explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

@@

In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously: the uniqueness of synthesised types and the decidability of bidirectional type synthesis for mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
For instance, we briefly discuss how the theory can be extended to handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed~\cite{Chen2024a} with \Agda but translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, with no gap in between.

### 2024-01-12 tex/ESOP24/S3-defs.tex
[ tex ] Eliminate most of hyphens the end of every line

\begin{example}
  Observe that $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
  \vspace{2em}
  \begin{equation*}
    \frac{\Gamma, \isTerm{x} : \eqnmarkbox[blue]{ext1}{A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
    \qquad\text{and}\qquad
    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
  \end{equation*}
  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
  \vspace{.5em}
  \\
  if the empty context $\cdot$ is added verbosely, so they can be specified by arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$, respectively, with $\Type_{\Sigma_{\bto}}(A, B)$ as types.
\end{example}

@@

\begin{defn}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
  \begin{inlineenum}
    \item variables in $V$,
    \item annotations $\isTerm{t \annotatecolon A}$ for some raw term $t$ in $V$ and a type $A$, and
    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
  \end{inlineenum}
  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
\end{defn}

@@

\begin{defn}\label{def:typing-derivations}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of 
  \begin{enumerate}
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotatecolon A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for an operation $\bop$ if there are $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
\end{defn}

@@

\begin{example}\label{ex:signature-simply-typed-lambda}
Having established generic definitions, we can now specify simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of 
\[
  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
  \quad\text{and}\quad
  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
\]
\end{example}
More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.

>>>>

\begin{example}
  Observe that the $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
  \vspace{2em}
  \begin{equation*}
    \frac{\Gamma, \isTerm{x} : \eqnmarkbox[blue]{ext1}{A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
    \qquad\text{and}\qquad
    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
  \end{equation*}
  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
  \vspace{.5em}
  \\
  if the empty context $\cdot$ is added verbosely, so they can be specified by arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$, respectively, with $\Type_{\Sigma_{\bto}}(A, B)$ as types.
\end{example}

@@

\begin{defn}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
  \begin{inlineenum}
    \item (indices of) variables in $V$,
    \item annotations $\isTerm{t \annotatecolon A}$ for some raw term $t$ in $V$ and a type $A$, and
    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
  \end{inlineenum}
  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
\end{defn}

@@

\begin{defn}\label{def:typing-derivations}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of 
  \begin{enumerate}
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotatecolon A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for some operation $\bop$ if there exist $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
\end{defn}

@@

\begin{example}\label{ex:signature-simply-typed-lambda}
Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of 
\[
  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
  \quad\text{and}\quad
  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
\]
\end{example}
More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.

### 2024-01-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Eliminate most of hyphens the end of every line

\begin{proposition}[Annotatability]\label{thm:annotatability}
If $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotatecolon A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotatecolon A) \sqsupseteq (\isTerm{u} \annotatecolon A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

On the other hand, when using a bidirectional type synthesiser to implement a type synthesiser, for example in \cref{thm:implementation}, if the bidirectional type synthesiser concludes that there does not exist a bidirectional typing derivation, we use the contrapositive form of completeness to establish that such an ordinary typing derivation does not exist either.
Now, annotatability is a kind of completeness because (roughly speaking) it turns an ordinary typing derivation bidirectional.
Therefore it is conceivable that we could use annotatability in place of completeness in the proof of \cref{thm:implementation}.
However, in the contrapositive form of annotatability, the antecedent is `there does not exist $t'$ that is more annotated than $t$ and has a bidirectional typing derivation', which is more complex than the bidirectional type synthesiser would have to produce.
Annotatability also does not help the user to deal with missing annotations: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that our theory offers simpler and more useful alternatives than the notion of annotatability.

>>>>

\begin{proposition}[Annotatability]\label{thm:annotatability}
If $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation given by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotatecolon A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotatecolon A) \sqsupseteq (\isTerm{u} \annotatecolon A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

On the other hand, when using a bidirectional type synthesiser to implement a type synthesiser, for example in \cref{thm:implementation}, if the bidirectional type synthesiser concludes that there does not exist a bidirectional typing derivation, we use the contrapositive form of completeness to establish that such an ordinary typing derivation does not exist either.
Now, annotatability is a kind of completeness because (roughly speaking) it turns an ordinary typing derivation bidirectional.
Hence, it is conceivable that we could use annotatability in place of completeness in the proof of \cref{thm:implementation}.
However, in the contrapositive form of annotatability, the antecedent is `there does not exist $t'$ that is more annotated than $t$ and has a bidirectional typing derivation', which is more complex than the bidirectional type synthesiser would have to produce.
Annotatability also does not help the user to deal with missing annotations: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that our theory offers simpler and more useful alternatives than the notion of annotatability.

### 2024-01-12 tex/ESOP24/S8-future.tex
[ tex ] Eliminate most of hyphens the end of every line

The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
Many general definitions and frameworks for defining logics and type theories have been proposed, and we classify them into two groups by where signatures reside: the meta level or the object level of a meta-language:
\begin{enumerate}
  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
\end{enumerate}

@@

For the second group, theories developed in set theory can often be reformulated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.

@@

To explore general settings, an extreme possibility is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.

>>>>

The idea of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
Many general definitions and frameworks for defining logics and type theories have been proposed and can be classified into two groups by where signatures reside: the meta level or the object level of a meta-language:
\begin{enumerate}
  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
\end{enumerate}

@@

For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.

@@

To explore general settings, an extreme choice is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
